COVID-19 Vaccine Acceptance and Hesitancy in Low and Middle Income Countries, and Implications for Messaging

### 2021-05-04 4_paper_supplementary_files/temp_download.Rmd
update files

<!-- COVID-19 Vaccine Acceptance and Hesitancy in Low and Middle Income Countries, and Implications for Messaging -->

@@

@@

We analyze COVID-19 vaccine acceptance across 15 survey samples covering ten low- and middle-income countries (LMICs) in Asia, Africa, and South America, and two higher income countries (Russia and the United States). Standardized survey responses were collected from `r nice(nrow(df))` individuals. We find generally high willingness to take a COVID-19 vaccine in LMIC samples (80% on average), considerably higher than in the United States (65%) and Russia (30%). Vaccine acceptance was primarily explained by an interest in personal protection against the disease, while concern about side effects was the most commonly expressed reason for reluctance. Health workers were the most trusted sources of information about COVID-19 vaccines. Our findings suggest that prioritizing efficient and equitable vaccine distribution to LMICs should yield high returns in promoting global immunization. Interventions to promote vaccination in these contexts should focus on translating intentions into uptake and involve trusted health workers.

@@

A safe and effective vaccine against COVID-19 is a critical tool to control the pandemic. As of March 2, 2021, there were 76 vaccines in clinical development and 16 had advanced to Stage 3 clinical trials [@who1]. Following clinical trials, several vaccines have been approved in multiple countries and are being distributed across the globe. At present, global vaccine distribution remains highly unequal, with much of the current supply directed toward high-income countries [@eiu].

While effective and equitable distribution of the vaccines is a key policy priority, ensuring the population’s acceptance is equally important. Acceptance of vaccines and trust in the institutions that administer them are likely key determinants of the success of any vaccination campaign [@defigueiredo2020lancet]. Several studies investigate high-income country residents’ willingness to take a potential COVID-19 vaccine [@boyon2020ipsos; @Malik2020]. Little is known about acceptance rates in low- and middle-income countries (LMICs), however, where the majority of the world’s population resides.

Acceptance of childhood vaccination for common diseases ---such as measles (MCV), Bacille Calmette-Guérin (BCG) and diphtheria, tetanus, and pertussis (DTP)---is generally high in LMICs, providing reasons for optimism about future uptake of COVID-19 vaccines. Table \ref{tab:otherv} summarizes general vaccine acceptance and uptake of common childhood vaccines for the countries included in our study. Still, existing studies on COVID-19 vaccine acceptance document large variation, both across and within countries, including in settings where overall vaccination rates are high [@boyon2020ipsos]. These studies highlight concerns about COVID-19 vaccine safety, and particularly concerns about the speed of vaccine development, as reasons for hesitancy in higher income settings. Similar concerns may apply in LMICs.

Additional reasons for hesitancy may feature more prominently in LMIC settings. Reported COVID-19 mortality rates have consistently been lower in LMICs relative to higher income countries [@mukherjee]. If individuals in LMICs feel the risk of disease is less serious, they may be less inclined to accept perceived risks associated with taking a recently-developed vaccine. Previous studies have also highlighted factors such as concerns about healthcare quality [@christensen2020building], negative historical experiences [@Lowes2018], weak support from traditional leaders [@Jegede2007], and mistrust in government [@BLAIR201789] as barriers to healthcare utilization in LMICs.

Understanding factors that may lead people in LMICs to reject COVID-19 vaccination is of global concern, since a lag in vaccination in the developing world could facilitate the spread of new variants of the virus to other countries.

In order to effectively promote vaccination against COVID-19 and devise messaging strategies, we need to know if people are willing to take COVID-19 vaccines, the reasons why they are willing or unwilling to do so, and the factors influencing their decision-making. For this purpose, we developed and deployed a common set of questions across 15 studies in ten LMICs in Africa (Burkina Faso, Mozambique, Nigeria, Rwanda, Sierra Leone, Uganda), Asia (India, Nepal, Pakistan), and Latin America (Colombia). We compare these findings in LMICs to those from two higher income countries (Russia and the United States).

@@

Our main results are shown in Figure \@ref(fig:mainfigure1) and reproduced as Table \ref{tab:maintabledis} in [Appendix A](#appendixd). The first column provides overall acceptance rates in each study, while the remaining columns show acceptance rates disaggregated by respondent characteristics. The “All LMICs” row reports averages for LMIC countries only (and so excludes Russia and the US).

We document meaningful variation in vaccine acceptance across and within LMICs, but generally high levels of acceptance in LMICs overall. The average acceptance across studies is `r ans_mean`% (95% CI `r ans_mean_low`--`r ans_mean_high`), with a median of `r ans_median`, a range of `r ans_max - ans_min` percent points and an interquartile range of `r ans_iqr`. Our estimate of $\tau^2$ is `r  round(mp_tau, 3)` which implies a standard deviation over country averages of  `r round(mp_tau^.5, 3)`.

The acceptance rate in every LMIC sample is higher than in USA (`r usa_ans %>% pull(estimate)`%, `r usa_ans %>% pull(conf.low)`--`r usa_ans %>% pull(conf.high)`) and Russia (`r rus_ans %>% pull(estimate)`%, `r rus_ans %>% pull(conf.low)`--`r rus_ans %>% pull(conf.high)`).

We find limited evidence of variation across demographic subgroups in LMICs, as shown in Table \ref{tab:dmeans}. Women were generally less willing to accept the vaccine (average difference about `r dif_gender` points, significant at $p<.01$). Younger respondents (defined as aged <55 given younger-skewing populations in LMICs) were marginally more willing to take the vaccine, but this difference was not statistically significant. Less educated people were on average more willing to take the vaccine in LMICs, but this difference was not significant.

To better understand the reasoning behind vaccine acceptance, we asked those who were willing to take the vaccine why they would take it. We summarize in Table \ref{tab:yes}, with more details in Table \ref{tab:yesall}.

The most common reason given for vaccine acceptance was personal protection against COVID-19 infection. The average across LMICs was `r yes_all$estimate`% (`r yes_all$conf.low`--`r yes_all$conf.high`). In every individual study, it ranked as the first reason. In distant second place, LMIC respondents reported willingness to take the COVID-19 vaccine in order to protect their families. The average across LMICs was `r yes_all_2$estimate`% (`r yes_all_2$conf.low`--`r yes_all_2$conf.high`). In comparison to self-protection, protecting the community did not feature prominently among the stated reasons.

Self-protection also ranked as the most commonly expressed reason for taking the vaccine in Russia (`r yes_rus$estimate`%, `r yes_rus$conf.low`--`r yes_rus$conf.high`) and the USA (`r yes_usa$estimate`%, `r yes_usa$conf.low`--`r yes_usa$conf.high`).

This evidence contrasts with appeals to altruistic behavior and prosocial motivations in order to promote vaccine acceptance [@chou2020considering]. The risks and benefits to personal well-being feature much more prominently in people’s stated reasons for vaccine acceptance.

Figure \@ref(fig:fig2paper) summarizes the reasons given among respondents who said they were not willing to take a Covid vaccine.

The most common reason expressed for reluctance to take the vaccine in LMIC studies was  concern about side effects. For studies  `r side_top$group[1]` (`r side_top$estimate[1]`% `r side_top$conf.low[1]`--`r side_top$conf.high[1]`), `r side_top$group[2]`  (`r side_top$estimate[2]`%, `r side_top$conf.low[2]`--`r side_top$conf.high[2]`), `r side_top$group[3]` (`r side_top$estimate[3]`% `r side_top$conf.low[3]`--`r side_top$conf.high[3]`) and `r side_top$group[4]` (`r side_top$estimate[4]`% `r side_top$conf.low[4]`--`r side_top$conf.high[4]`), more than half of those respondents unwilling to take the vaccine mentioned this reason. Respondents in Russia (`r side_rus$estimate`%, `r side_rus$conf.low`--`r side_rus$conf.high`) and the USA (`r side_usa$estimate`%, `r side_usa$conf.low`--`r side_usa$conf.high`), reported high levels of this same concern.

While serious adverse events that are life-threatening or require hospitalization are very rare, with only .6\% of respondents reporting at least one side effect in the Pfizer vaccine trial [@cdcadverse], one potential explanation for the outsized concern about side effects could be the lack of widespread information about features of the vaccine at the time of data collection. Media coverage of the few cases of serious adverse events and spread of fake news may contribute as well [@stein2017golden].

Concerns about side effects could also be due to a concern about mild side effects from experiences with other vaccines. In the case of available COVID-19 vaccines, we now know that mild side effects are common but transient. These include fatigue, muscle pain, joint pain and headache, which were severe in fewer than 10\% of people in the clinical trials of tens of thousands. Severe fever occurred in fewer than 2\% of them. [@wadman2020public]

Allergic reactions from the COVID-19 vaccine seem to be extremely rare. Data from trials of the Pfizer-BioNTech vaccines shows that anaphylaxis after reported administration occurs at a rate of 11.1 cases per million vaccine doses administered [@cdcallergies].  Our data reflects this --  no more than 6% of respondents expressed concern about allergies in any of our LMIC studies.

Other concerns that make many respondents unwilling to take the vaccine could be countered by accurately presenting the scientific data to the public. Studies `r effective_top$group[1]` (`r effective_top$estimate[1]`%, `r effective_top$conf.low[1]`--`r effective_top$conf.high[1]`), `r effective_top$group[2]`  (`r effective_top$estimate[2]`%, `r effective_top$conf.low[2]`--`r effective_top$conf.high[2]`) and `r effective_top$group[3]` (`r effective_top$estimate[3]`%, `r effective_top$conf.low[3]`--`r effective_top$conf.high[3]`) showed relatively high levels of skepticism about vaccine effectiveness. This is also true for respondents in Russia (`r effective_rus$estimate`%, `r effective_rus$conf.low`--`r effective_rus$conf.high`) and the USA (`r effective_usa$estimate`%, `r effective_usa$conf.low`--`r effective_usa$conf.high`). Recent clinical trials reveal very high rates of vaccine efficacy [@baden2021efficacy; @polack2020safety], and clearly communicating these results to the public is a high priority.  In contrast, conspiracy theories were rarely mentioned by respondents in any of our study samples, in spite of widespread popular discourse about anti-vaxxer movements and theories in higher-income countries [@loomba_measuring_2021].

Finally, respondents in some studies downplayed the seriousness of this disease, and listed this as a reason not to be vaccinated. Studies `r ill_top$group[1]` (`r ill_top$estimate[1]`% `r ill_top$conf.low[1]`--`r ill_top$conf.high[1]`), `r ill_top$group[2]`  (`r ill_top$estimate[2]`%, `r ill_top$conf.low[2]`--`r ill_top$conf.high[2]`) and `r ill_top$group[3]` (`r ill_top$estimate[3]`% `r ill_top$conf.low[3]`--`r ill_top$conf.high[3]`) reported high rates of respondents citing lack of concern as a reason they were hesitant.

The analysis above identifies the nature of the information gaps that any vaccine messaging should target, while in Figure \@ref(fig:fig3paper) we try to identify the actors who are best placed to deliver those messages. We asked respondents about their most trusted source of information when deciding whether to take the vaccine, because these sources are vital to disease control strategies during public health emergencies [@siegrist2014role]. Results from Figure \@ref(fig:fig3paper) are reproduced as Table \ref{tab:trust} in [Appendix A](#appendixd).

We find striking consistency across countries. In all but one study, respondents identified the health system as the most trustworthy source to help them decide whether or not to take the COVID-19 vaccine (with the exception of Rwanda, where the government in general was identified as the most trusted source, with the health system a close second). Family and friends were the next most important reference points in most samples. Across samples, women were 3 percentage points more likely to rely on family and friends than male respondents, though this difference was not significant at conventional levels (Figure \@ref(fig:genderhist) in [Appendix D](#appendixd)). By contrast, endorsements by religious leaders or celebrity figures were not seen as important sources of influence in any sample other than Nepal.

@@

To our knowledge, this is the first study documenting rates of expressed COVID-19 vaccine acceptance and hesitancy in a large set of LMICs. Our findings show variable but broadly high levels of prospective COVID-19 vaccine acceptance across LMICs using data from `r nice(nrow(filter(df, country!="USA" | country!="Russia")))` respondents in `r length(unique(df$country))-2` original household surveys from Africa (Burkina Faso, Mozambique, Nigeria, Rwanda, Sierra Leone, Uganda), Asia (Bangladesh, India, Nepal, Pakistan), and Latin America (Colombia). Acceptance across LMIC averages `r ans_mean`, ranging between  `r ans_min` and `r ans_max`. We document considerably lower levels of acceptance in Russia and the United States.

Patterns of COVID-19 vaccine hesitancy are not well predicted by existing measures of concerns about the safety of other vaccines (e.g., the Wellcome Global Monitor shown in Table \ref{tab:otherv}). Compared with other vaccines, COVID-19 vaccine acceptance is lower and more variable across LMIC samples. This suggests that concerns may apply specifically to COVID-19 vaccines rather than to vaccination more broadly.

Our study also documents reasons why respondents express intentions to take (or not take) a COVID-19 vaccine. The main reason expressed for willingness to take such a vaccine was to protect oneself. The most common reasons offered by those unwilling to take the vaccine were concerns about safety (side effects) and efficacy. Across all contexts, health care workers were the most trusted source of information  about vaccines.

Our study samples offer an important window into the motivations underlying COVID-19 vaccine acceptance in LMICs, but our data are not fully nationally representative. Random digit dial samples and follow-up phone surveys, while necessary during a global pandemic, do not include individuals who reside outside coverage areas, who do not own or cannot operate cell phones, or who choose not to respond to telephone surveys. Care should also be taken in any attempt to extrapolate to the population level from the samples representative of narrow subpopulations.

If intentions reported in our LMIC samples translated into actual vaccination uptake, the rates would far exceed the range of what would be required for COVID-19 herd immunity (40-67% in recent estimates [@britten2020; @haley2020]). However, reported intent may not materialize into actual vaccine adoption [@mceachanetal2011]. The high salience of COVID-19 due to extensive media coverage and government mitigation efforts and excitement around vaccine release may have increased reported intention [@chen1996]. Results from  the first COVID-19 vaccine Phase 3 clinical trial were announced before some surveys were fielded; during others, subsequent approvals were granted. The fast-moving information environment may change people’s perceptions about vaccines by the time they are widely available in LMICs.

Nonetheless, our findings provide some specific guidance on how to design messaging to boost COVID-19 vaccine acceptance and uptake in LMICs. Our data have implications for both what the content of the message should be, and who should deliver the message.

First, high levels of trust in the advice of health workers and governments on COVID-19 vaccine decision-making suggest that social and behavioral change communication (SBCC) strategies that engage local health workers may be particularly effective tools to encourage timely and complete vaccine uptake, and to combat remaining vaccine hesitancy. The literature has  explored messaging strategies to promote welfare-improving behaviors, with considerable attention paid to celebrity endorsements [@alatas2019celebrities]. Our data strongly support the view [@bokemper2021timing] that those with the most relevant expertise - as opposed to celebrities or general opinion leaders - are most trusted on this specific topic and are therefore best positioned to deliver the message.

Second, the average COVID-19 vaccine acceptance rate across our LMIC samples is high, approximately `r ans_mean`\%. Given such positive intentions, there may be high returns to investing in straightforward “last-mile” nudges that help citizens convert intentions into actions. Reminder messages from healthcare providers and messages alerting patients that vaccines have been reserved for them at an upcoming appointment may provide a low-cost encouragement to initiate and complete two-dose COVID-19 vaccinations, as was found in two recent large-scale studies in the United States [@milkmanetal2021a; @milkmanetal2021b]. Similarly, childhood vaccination reminders plus cash incentives in Kenya substantially increased full immunization [@gibsonetal2017], and cash and in-kind incentive programs in Nigeria and India have also proven effective [@idinsight; @Banerjee2010].

Third, high coverage rates of existing vaccines, coupled with respondents’ reliance on friends and family as information sources, suggest that the general pro-vaccination stance of many citizens could be leveraged to convert intent to uptake. Social learning strategies and norm-setting are powerful drivers of information diffusion and behavior change in many related sectors [@beaman2021can]. Social signalling of positive attitudes toward COVID-19 vaccines may also help shift social norms toward even greater immunization acceptance and two-dose completion in the community at large [@karing2018social].

Finally, our findings offer guidance on the specific content of vaccine messaging that is likely to be most persuasive. Messaging should highlight the high efficacy rates of the COVID-19 vaccines currently on the market in reducing or eliminating disease, hospitalizations and death. Alluding to clinical data that addresses people’s concerns about potential side effects should be prioritized. Messaging should also emphasize the direct protective benefits of the vaccine to the adopter.

>>>>

<!-- KEEP ALL REVISIONS IN SUGGESTING MODE FOR VERSION CONTROL - DO NOT ACCEPT CHANGES -->
<!-- COVID-19 Vaccine Acceptance and Hesitancy in Low and Middle Income Countries, and Implications for Messaging -->

@@

```{r, include=FALSE} 
empty = " " 
```
`r empty`

@@

We analyze COVID-19 vaccine acceptance across 15 survey samples covering ten low- and middle-income countries (LMICs) in Asia, Africa, and South America, Russia (an upper-middle-income country), and the United States, using survey responses from  `r nice(nrow(df))` individuals. We find considerably higher willingness to take a COVID-19 vaccine in LMIC samples (80% on average) compared to the United States (65%) and Russia (30%). Vaccine acceptance was primarily explained by an interest in personal protection against COVID-19, while concern about side effects was the most commonly expressed reason for reluctance. Health workers were the most trusted sources of information about COVID-19 vaccines. Our findings suggest that prioritizing vaccine distribution to LMICs should yield high returns in promoting global immunization coverage, and that vaccination campaigns in these countries should focus on translating acceptance into uptake. Messaging highlighting vaccine efficacy and safety, delivered by healthcare workers, may be most effective in addressing remaining hesitancy.

@@

A safe and effective vaccine against COVID-19 is a critical tool to control the pandemic. As of April 27, 2021, 22 vaccines had advanced to Stage 3 clinical trials [@who1] and more than a dozen vaccines had been approved in multiple countries[@track]. For example, as of April 27, 2021, Comirnaty (BioNTech/Pfizer) had been approved in  more than 80 countries, while the AstraZeneca vaccine had the most country authorizations at 92. [@track] At present, however, global vaccine distribution remains highly unequal, with much of the current supply directed toward high-income countries [@wouters2021challenges].

While effective and equitable distribution of COVID-19 vaccines is a key policy priority, ensuring the population’s acceptance is equally important. Trust in vaccines and the institutions that administer them are key determinants of the success of any vaccination campaign [@defigueiredo2020lancet]. Several studies have investigated willingness to take a potential COVID-19 vaccine in high-income countries [@boyon2020ipsos; @Malik2020; @kreps2020factors; @shekhar2021covid; @dror2020vaccine; @fisher2020attitudes], and some studies have included middle-income countries [@lazarus2020nature; @wouters2021challenges]. Less is known, however, about vaccine acceptance in low-income countries where large-scale vaccination  has yet to begin. Understanding the drivers of COVID-19 vaccine acceptance is of global concern, since a lag in vaccination in any country may result in the emergence and spread of new variants that can overcome vaccine- and prior disease-conferred immunity [@ong2021lack; @lancettask].

Our study complements the emerging global picture of COVID-19 vaccine acceptance by focusing primarily on lower income countries. We construct a sample of low- and middle-income countries (LMICs) with wide geographic coverage across Africa, Asia and Latin America. We move beyond documenting vaccine acceptance rates and analyze data on the reasons for acceptance and hesitancy, which is critical for informing the design of effective vaccine distribution and messaging.

Acceptance of childhood vaccination for common diseases---such as measles (MCV), Bacille Calmette-Guérin (BCG) and diphtheria, tetanus, and pertussis (DTP)---is generally high in LMICs, providing cause for optimism about the prospects for COVID-19 vaccine uptake. Table \ref{tab:otherv} summarizes general vaccine acceptance and coverage rates of childhood vaccines in 2018, prior to the current pandemic,  for the countries included in our study. Agreement on the importance of childhood vaccinations is markedly higher in these LMICs compared to Russia and the United States. Still, existing studies on COVID-19 vaccine acceptance document substantial variation, both across and within countries, including in settings with high background levels of other vaccinations [@defigueiredo2020lancet; @lazarus2020nature; @wouters2021challenges].

This literature cites concerns about COVID-19 vaccine safety, including concerns about the rapid pace of vaccine development, as reasons for hesitancy in higher-income settings [@boyon2020ipsos; @wouters2021challenges], but other reasons  may feature more prominently in LMICs. For instance, reported COVID-19 mortality rates have been consistently lower in LMICs relative to higher income countries [@mukherjee]. If individuals feel the risk of disease is less serious, they may be less willing to accept any perceived risks of vaccination [@brewer2007meta]. Previous studies of healthcare utilization in LMICs have also highlighted factors such as healthcare quality concerns [@christensen2020building], negative historical experiences involving foreign actors  [@Lowes2018; @martinez2021vaccines], weak support from traditional leaders [@Jegede2007], and mistrust in government [@BLAIR201789] as barriers to uptake, which could apply to COVID-19 vaccination as well.

To promote vaccination against COVID-19 and develop effective messaging strategies, we need to know whether people are willing to take COVID-19 vaccines, the reasons why they are willing or unwilling to do so, and the most trusted  sources in their decision-making. Our study investigates these questions using a common set of survey items deployed across 13 studies in Africa, South Asia, and Latin America: seven in low-income countries (Burkina Faso, Mozambique, Rwanda, Sierra Leone, Uganda), five in lower-middle-income countries (India, Nepal, Nigeria, Pakistan), and one in an upper-middle-income country (Colombia). We compare these findings to those from two countries at the forefront of vaccine research and development, Russia and the United States. 
To select studies to include in our sample, we conducted an internal search within Innovations for Poverty Action (IPA), the International Growth Center (IGC), and the Berlin Social Science Center (WZB) for projects with plans to collect survey data in the second half of 2020. Study PIs agreed to include a set of common questions about COVID-19 vaccine attitudes. This strategy was guided by the need to collect information quickly and cost effectively using a survey modality (phone) that was both safe given pandemic conditions and appropriate for contexts with limited internet coverage. The final set of samples included in our study therefore reflects populations that fall under the current research priorities at IGC, IPA and WZB and, in the case of IPA and IGC, donors that prioritize working in the Global South.

@@

Our main results are shown in Figure \@ref(fig:mainfigure1) and reproduced as [Appendix A](#appendixd) Table \ref{tab:maintabledis}. The first column provides overall acceptance rates in each study, while the remaining columns disaggregate acceptance rates by respondent characteristics. The “All LMICs” row reports averages for LMIC countries and excludes Russia and the USA.

We document meaningful variation in vaccine acceptance across and within LMICs, but generally high levels of acceptance in LMICs overall. The average acceptance rate across studies was `r ans_mean`% (95% CI `r ans_mean_low`--`r ans_mean_high`%), with a median of `r ans_median`, a range of `r ans_max - ans_min` percent points and an interquartile range of `r ans_iqr`. Our estimate of $\tau^2$ is `r  round(mp_tau, 3)` which implies a standard deviation over country averages of  `r round(mp_tau^.5, 3)`.

The acceptance rate in every LMIC sample was higher than in the USA (`r usa_ans %>% pull(estimate)`%, `r usa_ans %>% pull(conf.low)`--`r usa_ans %>% pull(conf.high)`%) and Russia (`r rus_ans %>% pull(estimate)`%, `r rus_ans %>% pull(conf.low)`--`r rus_ans %>% pull(conf.high)`%). Reported acceptance was lowest in `r bottom_all$group[1]` (`r bottom_all$estimate[1]`% `r bottom_all$conf.low[1]`--`r bottom_all$conf.high[1]`%) and `r bottom_all$group[2]`  (`r bottom_all$estimate[2]`%, `r bottom_all$conf.low[2]`--`r bottom_all$conf.high[2]`%). The case of Pakistan may be linked to negative historical experiences with foreign-led vaccination campaigns [@ali2019polio; @robbins2012cia]. This hesitancy may be particularly problematic, given the magnitude of the second wave in neighboring India just as vaccines are becoming available, with accelerating cases in Bangladesh, Pakistan and India threatening to overwhelm health infrastructure.

We find limited evidence of variation across demographic subgroups in our LMIC samples, as shown in [Appendix A](#appendixd) Table \ref{tab:dmeans}. Women were generally less willing to accept the vaccine (average difference about `r dif_gender` points, significant at $p<.01$). Respondents under age 24 and less educated respondents were marginally more willing to take the vaccine, but these differences were not statistically significant. The acceptance advantage among the less educated was most pronounced in Rwanda, Burkina Faso, and Uganda, and stands in contrast to the USA, where more educated respondents were significantly more willing to be vaccinated.

To better understand the reasoning behind vaccine acceptance, we asked those who were willing to take the vaccine why they would take it. We summarize these results in Table \ref{tab:yes}, with additional details in [Appendix A](#appendixd) Table \ref{tab:yesall}. The reason most commonly given for vaccine acceptance across samples was personal protection against COVID-19 infection. The average across the LMIC samples was `r yes_all$estimate`% (`r yes_all$conf.low`--`r yes_all$conf.high`%). In every individual study, this ranked as the most popular reason. In distant second place among LMIC respondents was family protection with an average of `r yes_all_2$estimate`% (`r yes_all_2$conf.low`--`r yes_all_2$conf.high`%). In comparison to protecting oneself and one’s family,  protecting one’s community did not feature prominently among stated reasons for acceptance. These reasons do not vary substantially by age group, as shown in Table \ref{tab:countrydiff}. Self-protection also ranked as the most common reason for taking the vaccine in Russia (`r yes_rus$estimate`%, `r yes_rus$conf.low`--`r yes_rus$conf.high`%) and the USA (`r yes_usa$estimate`%, `r yes_usa$conf.low`--`r yes_usa$conf.high`%).

Figure \@ref(fig:fig2paper) summarizes the reasons given by respondents who said they were not willing to take a COVID-19 vaccine. Concern about side effects was the most frequently expressed reason for reluctance in our  LMIC samples. This concern was particularly evident among African country samples. In studies  `r side_top$group[1]` (`r side_top$estimate[1]`% `r side_top$conf.low[1]`--`r side_top$conf.high[1]`%), `r side_top$group[2]`  (`r side_top$estimate[2]`%, `r side_top$conf.low[2]`--`r side_top$conf.high[2]`%), `r side_top$group[3]` (`r side_top$estimate[3]`% `r side_top$conf.low[3]`--`r side_top$conf.high[3]`%) and `r side_top$group[4]` (`r side_top$estimate[4]`% `r side_top$conf.low[4]`--`r side_top$conf.high[4]`%), more than half of respondents unwilling to take the vaccine cited worries about side effects.  Respondents in Russia (`r side_rus$estimate`%, `r side_rus$conf.low`--`r side_rus$conf.high`%) and even more in the USA (`r side_usa$estimate`%, `r side_usa$conf.low`--`r side_usa$conf.high`%), frequently reported this same concern.

Study samples `r effective_top$group[1]` (`r effective_top$estimate[1]`%, `r effective_top$conf.low[1]`--`r effective_top$conf.high[1]`%), `r effective_top$group[2]`  (`r effective_top$estimate[2]`%, `r effective_top$conf.low[2]`--`r effective_top$conf.high[2]`%) and `r effective_top$group[3]` (`r effective_top$estimate[3]`%, `r effective_top$conf.low[3]`--`r effective_top$conf.high[3]`%) showed relatively high levels of skepticism about vaccine effectiveness among hesitant respondents. This was also true in Russia (`r effective_rus$estimate`%, `r effective_rus$conf.low`--`r effective_rus$conf.high`%) and the USA (`r effective_usa$estimate`%, `r effective_usa$conf.low`--`r effective_usa$conf.high`%). In addition, hesitant respondents listed lack of concern about COVID-19 infection as a reason not to be vaccinated. Study samples `r ill_top$group[1]` (`r ill_top$estimate[1]`% `r ill_top$conf.low[1]`--`r ill_top$conf.high[1]`%), `r ill_top$group[2]`  (`r ill_top$estimate[2]`%, `r ill_top$conf.low[2]`--`r ill_top$conf.high[2]`%) and `r ill_top$group[3]` (`r ill_top$estimate[3]`% `r ill_top$conf.low[3]`--`r ill_top$conf.high[3]`%) reported high rates of this answer among hesitant respondents.

In Figure \@ref(fig:fig3paper) we report respondents’ most trusted source of guidance when deciding whether to take a COVID-19 vaccine. Results from Figure \@ref(fig:fig3paper) are reproduced as  [Appendix A](#appendixd) Table \ref{tab:trust}.  [Appendix B](#appendixb) Table \ref{tab:q4} presents a complete description of response recoding from individual studies.

We find striking consistency in most trusted sources across studies. In all but one sample, including those from Russia and the USA, respondents identified the health system as the most trustworthy source to help them decide whether to take the COVID-19 vaccine. The average across LMICs was `r trust_all$estimate`% (`r trust_all$conf.low`--`r trust_all$conf.high`%). Respondents in `r trust_top$group[1]` (`r trust_top$estimate[1]`%, `r trust_top$conf.low[1]`--`r trust_top$conf.high[1]`%), `r trust_top$group[2]`  (`r trust_top$estimate[2]`%, `r trust_top$conf.low[2]`--`r trust_top$conf.high[2]`%) and `r trust_top$group[3]` (`r trust_top$estimate[3]`%, `r trust_top$conf.low[3]`--`r trust_top$conf.high[3]`%) cited  health workers most often. Sierra Leone had the highest trust in health workers and the Ministry of Health, potentially due to investments in public health following the 2014-2015 Ebola epidemic [@deserrano].

The next most cited sources were family and/or friends in  `r trust_fam$group[1]` (`r trust_fam$estimate[1]`%, `r trust_fam$conf.low[1]`--`r trust_fam$conf.high[1]`%), `r trust_fam$group[2]`  (`r trust_fam$estimate[2]`%, `r trust_fam$conf.low[2]`--`r trust_fam$conf.high[2]`%), `r trust_fam$group[3]` (`r trust_fam$estimate[3]`%, `r trust_fam$conf.low[3]`--`r trust_fam$conf.high[3]`%) and `r trust_fam$group[4]` (`r trust_fam$estimate[4]`%, `r trust_fam$conf.low[4]`--`r trust_fam$conf.high[4]`%). Across the pooled samples, women were 3 percentage points more likely to rely on family and friends than male respondents, though this difference was not statistically significant (Figure \@ref(fig:genderhist) in [Appendix D](#appendixd)).

By contrast, in `r trust_gov$group[1]` (`r trust_gov$estimate[1]`%, `r trust_gov$conf.low[1]`--`r trust_gov$conf.high[1]`%), `r trust_gov$group[2]`  (`r trust_gov$estimate[2]`%, `r trust_gov$conf.low[2]`--`r trust_gov$conf.high[2]`%), `r trust_gov$group[3]` (`r trust_gov$estimate[3]`%, `r trust_gov$conf.low[3]`--`r trust_gov$conf.high[3]`%) and `r trust_gov$group[4]` (`r trust_gov$estimate[4]`%, `r trust_gov$conf.low[4]`--`r trust_gov$conf.high[4]`%), government was the second most frequently cited. Religious leaders and celebrities were not seen as the top sources of guidance by many respondents in any sample other than Nepal, where many respondents most trusted famous people (`r trust_npl$estimate`%, `r trust_npl$conf.low`--`r trust_npl$conf.high`%).

Finally, we highlight two idiosyncratic, yet frequently mentioned, trusted sources of information in deciding whether to take a COVID-19 vaccine. In Rwanda, 34% of respondents would most trust “themselves” for guidance, the most frequent response in this sample. In the USA, 14% of respondents cited Joe Biden, then president-elect and therefore excluded from the “government” category, as their most trusted source.

@@

<!-- Temp subheading: interpretation of results -->
To our knowledge, this is the first study focusing on COVID-19 vaccine attitudes in a large set of low-income and lower-middle-income countries. Our findings showed variable but broadly high levels of prospective COVID-19 vaccine acceptance across LMICs, using data from `r nice(nrow(filter(df, country!="USA" | country!="Russia")))` respondents in `r length(unique(df$country))-2` studies in ten LMICs in Africa, South Asia, and Latin America. Acceptance across these LMIC samples  averaged `r ans_mean`%, ranging between `r ans_min`% and `r ans_max`%. The two benchmark countries, Russia and the USA, demonstrated lower COVID-19 vaccine acceptance, consistent with relatively lower pre-pandemic vaccine confidence.

Many metrics and indices measure vaccine acceptance and hesitancy globally [@betsch2018beyond; @domek2018measuring; @shapiro2018vaccine; @gilkey2014vaccination,]. Our surveys used measures employed in other COVID-19 vaccine acceptance studies [@fisher2020attitudes;  @Malik2020; @wouters2021challenges; @lazarus2020nature; @kreps2020factors; @shekhar2021covid; @dror2020vaccine] and recommended by the WHO Data for Action guidance [@whodata], allowing for meaningful cross-study and cross-country comparisons. We measure trust in sources of information about COVID-19 vaccination using a measure similar to that used in the Vaccine Confidence Index (VCI), a widely used survey tool [@defigueiredo2020lancet].

Consistent with other studies, we find higher vaccine acceptance among men than women [@kreps2020factors; @shekhar2021covid; @dror2020vaccine;  @fisher2020attitudes; @wouters2021challenges]. In contrast to studies focused primarily on higher-income countries, we find no consistently significant differences with respect to age [@kreps2020factors; @dror2020vaccine] or education in our LMIC samples.

A key contribution of our study relative to the previous literature is its focus on the reasons why respondents express intentions to take (or refuse) a COVID-19 vaccine. Other work has highlighted appeals to altruistic behavior or other prosocial motivations to promote vaccine acceptance [@chou2020considering]. Yet we find that the potential risks and benefits to personal well-being feature much more prominently in our respondents’ reasoning, suggesting that appeals about personal protection may be more effective in LMICs.

The most commonly stated reason for vaccine refusal was concern about safety (side effects). Around 86% of our surveys were conducted as reports from Phase 2 and 3 clinical trial data were emerging for the earliest commercially available vaccines, but prior to the first Emergency Use Authorization of any vaccine (Pfizer-BioNTech approved by the USA on December 11, 2020). Early trial data showed that severe adverse effects were extremely rare [@mulligan2020phase; @logunov2021safety; @folegatti2020safety; @voysey2021safety; @baden2021efficacy; @polack2020safety], occurring in fewer than 10\% of people in clinical trials [@wadman2020public]. Our respondents’ outsized concern about side effects could reflect the rapid pace of vaccine development, [@callaway2020russia] and limited information available about potential COVID-19 vaccine safety at the time of data collection. These concerns could also reflect worries about mild, yet common and transient side effects, such as fatigue, muscle pain, joint pain and headache.

Prominent media coverage of adverse events may exacerbate concerns about side effects [@stein2017golden]. In particular, new information about rare but severe cases of thrombosis associated with the Astra-Zeneca vaccine that appeared after our data collection period could affect hesitancy levels. This is of particular relevance to LMICs, which are likely to rely on the Astra-Zeneca vaccine in their immunization campaigns, through initiatives such as COVAX.

Concerns about vaccine efficacy, averaging approximately `r filter(effective_top, group=="All") %>% pull(estimate)`% in the LMIC samples, may also reflect a lack of information about vaccines at the time of our surveys.  Strikingly, even in low-information environments, respondents in our samples rarely cited conspiracy theories about ulterior motives on the part of corporations, politicians or the pharmaceutical industry, despite widespread popular fears about these issues in higher-income countries [@loomba_measuring_2021].

<!-- Temp subheading: limitations -->
An important limitation of our study is that our samples are not fully nationally representative. Phone surveys, while necessary during a global pandemic, do not include individuals who reside outside coverage areas, lack access to a cell phone, or do not respond to calls. In addition, as shown in Table \ref{tab:sampling}, several studies focus on sub-populations of interest in pre-existing studies to which questions about COVID-19 vaccination were added. Particular care should be taken in any attempt to extrapolate to the national population level from these samples. Furthermore, we emphasize that our data are not representative of all LMICs. They represent a convenience sample of studies in countries in which our organizations could quickly mobilize coordinated data collection.
